\section{Group actions}
\secbegin{secGroupActions}

\todo{}

\begin{definition}
\label{defGroupAction}
Let $G$ be a group and $X$ be a set. A (\textbf{left}) \textbf{action} of $G$ on $X$ is a function
\[ \alpha : G \times X \to X \]
such that, writing $g \cdot x = \alpha(g,x)$ for all $g \in G$ and $x \in X$, we have:
\begin{enumerate}[(i)]
\item $e \cdot x = x$ for all $x \in X$; and
\item $(gh) \cdot x = g \cdot (h \cdot x)$ for all $g,h \in G$ and all $x \in X$.
\end{enumerate}
\end{definition}

\todo{}

\begin{theorem}
\label{thmGroupActionsHomomorphismsToSymmetricGroup}
Let $G$ be a group and let $X$ be a set. Actions of $G$ on $X$ correspond with group homomorphisms $G \to \mathrm{Sym}(X)$.
\end{theorem}

\begin{cproof}
Given an action $\alpha : G \times X \to X$, define $\alpha_g : X \to X$ by $\alpha_g(x) = g \cdot x$ for all $g \in G$ and all $x \in X$. Then $\alpha_g \in \mathrm{Sym}(X)$ for all $g \in G$---indeed, for all $x \in X$ we have
\[ \alpha_{g^{-1}}(\alpha_g(x)) = g^{-1} \cdot (g \cdot x) = (g^{-1}g) \cdot x = e \cdot x = x \]
and likewise $\alpha_g(\alpha_{g^{-1}}(x)) = x$ for all $x \in X$, so that $\alpha_g$ is a bijection with inverse $(\alpha_g)^{-1} = \alpha_{g^{-1}}$.

So define $\sigma_{\alpha} : G \to \mathrm{Sym}(X)$ by $\sigma_{\alpha}(g) = \alpha_g$. Then given $g,h \in G$ and $x \in X$, we have
\[ \sigma_{\alpha}(gh)(x) = \alpha_{gh}(x) = (gh) \cdot x = g \cdot (h \cdot x) = \alpha_g(\alpha_h(x)) = (\sigma_{\alpha}(g) \sigma_{\alpha}(h)) (x) \]
so that $\sigma_{\alpha}(gh) = \sigma_{\alpha}(g)\sigma_{\alpha}(h)$ for all $g,h \in G$. Hence $\sigma_{\alpha}$ is a group homomorphism.

It remains to prove that the assignment $\alpha \mapsto \sigma_{\alpha}$ defines a bijection from the set of actions of $G$ on $X$ to the set of homomorphisms $G \to \mathrm{Sym}(X)$.

\begin{itemize}
\item \textbf{Injectivity.} Let $\alpha$ and $\beta$ be actions of $G$ on $X$ and assume $\sigma_{\alpha} = \sigma_{\beta}$. To simplify notation, write $\alpha(g,x) = g \cdot x$ and $\beta(g,x) = g \star x$ for all $g \in G$ and $x \in X$. Then for all $g \in G$ and $x \in X$, we have
\[ g \cdot x = \alpha_g(x) = (\sigma_{\alpha}(g))(x) = (\sigma_{\beta}(g))(x) = \beta_g(x) = g \star x \]
and so $\alpha = \beta$, as required.

\item \textbf{Surjectivity.} Let $\sigma : G \to \mathrm{Sym}(X)$ be a homomorphism, and define $\alpha : G \times X \to X$ by $\alpha(g,x) = \sigma(g)(x)$. Then writing $g \cdot x = \alpha(g,x)$ for all $g \in G$ and $x \in X$, we have
\begin{enumerate}
\item $e \cdot x = \sigma(e)(x) = \mathrm{id}_X(x) = x$; and
\item $(gh) \cdot x = \sigma(gh)(x) = (\sigma(g)\sigma(h))(x) = \sigma(g)(h \cdot x) = g \cdot (h \cdot x)$.
\end{enumerate}
So $\alpha$ is an action. Moreover for all $g \in G$ and $x \in X$ we have
\[ \sigma_{\alpha}(g)(x) = \alpha_g(x) = g \cdot x = \sigma(g)(x) \]
and so $\sigma = \sigma_{\alpha}$, as required.
\end{itemize}

Therefore the assignment $\alpha \mapsto \sigma_{\alpha}$ establishes a bijection between the set of actions of $G$ on $X$ and the set of group homomorphisms $G \to \mathrm{Sym}(X)$.
\end{cproof}

\todo{}

\subsection*{Orbits and stabilisers}

\todo{}

\begin{definition}
\label{defOrbit}
Let $G$ be a group acting on a set $X$. The \textbf{orbit} of an element $x \in X$ under the action of $G$ is the subset $G \cdot x \subseteq X$ defined by
\[ G \cdot x = \{ g \cdot x \mid x \in X \} \]
\end{definition}

\todo{}

\begin{definition}
\label{defStabiliser}
Let $G$ be a group acting on a set $X$. The \textbf{stabiliser} of an element $x \in X$ under the action of $G$ is the subset $G_x \subseteq G$ defined by
\[ G_x = \{ g \in G \mid g \cdot x = x \} \]
In fact $G_x$ is a subgroup of $G$---see \Cref{exStabiliserIsSubgroup}.
\end{definition}

\begin{exercise}
Let $G$ be a group acting on a set $X$. Prove that for all $x \in X$, the stabiliser $G_x$ is a subgroup of $G$.
\hintlabel{exStabiliserIsSubgroup}{%
Given $g,h \in G_x$, prove that $(gh^{-1}) \cdot x = x$ and apply \Cref{strCriterionForSubgroup}.
}
\end{exercise}

\todo{}

\begin{theorem}[Orbit--stabiliser theorem]
\label{thmOrbitStabiliser}
Let $G$ be a finite group acting on a set $X$. Then $|G| = |G \cdot x| |G_x|$ for all $x \in X$.
\end{theorem}

\begin{cproof}
Fix $x \in X$. We'll prove that $|G \cdot x| = [G : G_x]$; the result is then immediate by Lagrange's theorem \Cref{thmLagrange}.

So define $f : G/G_x \to G \cdot x$ by $f(aG_x) = a \cdot x$ for all $a \in G$. Then
\begin{itemize}
\item \textbf{$f$ is well-defined.} To see this, let $a,b \in G$ and assume that $aG_x = bG_x$. Then $b^{-1}a \in G_x$, so that
\[ a \cdot x = (bb^{-1}a) \cdot x = b \cdot ((b^{-1}a) \cdot x) = b \cdot x \]
and so $f(aG_x) = f(bG_x)$, as required.

\item \textbf{$f$ is injective.} Let $aG_x, bG_x \in G/G_x$ and assume that $a \cdot x = b \cdot x$. Then
\[ (b^{-1}a) \cdot x = b^{-1} \cdot (a \cdot x) = b^{-1} \cdot (b \cdot x) = (b^{-1}b) \cdot x = e \cdot x = x \]
so that $b^{-1}a \in G_x$, and hence $aG_x = bG_x$.

\item \textbf{$f$ is surjective.} Let $y \in G \cdot x$. Then $y = a \cdot x = f(aG_x)$ for some $a \in G$, as required.
\end{itemize}

Since $f$ is a bijection, we have $|G \cdot x| = |G/G_x| = [G:G_x]$. Thus
\[ |G| = [G:G_x]|G_x| = |G \cdot x||G_x| \]
by Lagrange's theorem.
\end{cproof}

\todo{}